---
title: "24_11_27_isoform_filtering"
author: "L Botzenhardt"
date: "2024-11-27"
output: html_document
---

```{r setup, include=FALSE, eval=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(stringr)
library(dplyr)
library(tidyr)
library(Seurat)
```


# Filtering Proteomes 
## Aiptasia
The Proteomes (sources) were filtered to keep only the longest Isoforms, and had duplicated sequences removed. 
```{bash, eval=FALSE}
##this way seems to be the most concise
#cut out column 9, remove the first 4 lines, get only lines with protein_id,
#get the gene and protein ID using awk from column 9 of GTF

cut -f 9  aip_genomic_refseq.gtf | tail -n+5 | grep "protein_id" | \
awk -F';' '{for(i=1;i<=NF;i++) if ($i ~ /gene_id/ || $i ~ /protein_id/) printf $i; print ""}'| \ 
cut -d " " -f2,4 | sed 's/"//g' | uniq > ../02_output/04_aiptasia/aip_gene_protID.tab

#getting the protein lengths
seqtk comp aip_protein_refseq.faa | cut -f1,2 > ../02_output/04_aiptasia/aip_protein_lengths.tab
```

```{r}
#read in the data
aip_gene_protID <- read.table("../02_output/04_aiptasia/aip_gene_protID.tab", header = FALSE, sep = " ")
aip_gene_protID <- aip_gene_protID %>% 
  rename("gene_id" = V1, "protein_id" = V2)

aip_protein_lengths <- read.table("../02_output/04_aiptasia/aip_protein_lengths.tab", header = FALSE, sep = "\t")
aip_protein_lengths <- aip_protein_lengths %>% 
  rename("protein_id" = V1, "length" = V2)
```

## Merge the data
```{r, eval=FALSE}
#merge by protein_id  
aip_gene_prot_lengths <- aip_gene_protID %>% 
  left_join(aip_protein_lengths, by = "protein_id")
```

## checking amount of duplicates
```{r}
aip_longest_transcript_n_copies <- aip_gene_prot_lengths %>% 
  group_by(gene_id) %>% 
  filter(length == max(length)) %>% 
  summarize(n = n())
  
```
## taking a single duplicate per gene ID
```{r}
aip_long_transcript_distinct <- aip_gene_prot_lengths %>% 
  group_by(gene_id) %>% 
  filter(length == max(length)) %>% 
  distinct(gene_id, length, .keep_all = TRUE)

write.table(aip_long_transcript_distinct, "../02_output/04_aiptasia/aip_long_transcript_distinct.tab", sep = "\t", quote = FALSE, row.names = FALSE)
write.table(aip_long_transcript_distinct$protein_id, "../02_output/04_aiptasia/aip_long_transcript_distinct_protein.txt", sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)
```

```{bash, eval=FALSE}
seqtk subseq ../../00_data/aip_protein_refseq.faa aip_long_transcript_distinct_protein.txt > aip_longest_isoforms.fasta
```

## Nematostella
#### extract gene_ids for proteins from gtf file
take only column containing gene + Protein ID
remove header
filter for lines that contain protein_ids
take only those to entries per row

```{bash, eval=FALSE}
##approach using awk
$ cut -f 9 nvec_genomic_jaNemVect1_1.gtf | tail -n+5 | grep "protein_id" | awk -F';' '{for(i=1;i<=NF;i++) if ($i ~ /gene_id/ || $i ~ /protein_id/) printf $i";"; print ""}' | uniq | sed -e "s/;/\t/g" -e "s/\"//g" -e "s/protein_id//g" -e "s/gene_id//g" -e "s/\t$//g" -e "s/ //g" > ../02_output/03_nvec/nvec_gene_protID.tab

#get protein lengths
$ seqtk comp nvec_protein_jaNemVect1_1.faa | cut -f1,2 > ../02_output/03_nvec/nvec_protein_lengths.tab

```
### Read into R and rename
```{r, eval=FALSE}
#read in the data
nvec_gene_protID <- read.table("../02_output/03_nvec/nvec_gene_protID.tab", header = FALSE, sep = "\t")
nvec_gene_protID <- nvec_gene_protID %>% 
  rename("gene_id" = V1, "protein_id" = V2)

nvec_protein_lengths <- read.table("../02_output/03_nvec/nvec_protein_lengths.tab", header = FALSE, sep = "\t")
nvec_protein_lengths <- nvec_protein_lengths %>% 
  rename("protein_id" = V1, "length" = V2)
```
### Merge the data
```{r, eval=FALSE}
#merge by protein_id  
nvec_gene_prot_lengths <- nvec_gene_protID %>% 
  left_join(nvec_protein_lengths, by = "protein_id")

nvec_n_protein_duplicates <- nvec_gene_prot_lengths %>% 
  group_by(gene_id) %>% 
  filter(length == max(length)) %>% 
  summarize(n = n())

```

### Filter longest isoform
# Filtering Proteomes 
## Aiptasia
The Proteomes (sources) were filtered to keep only the longest Isoforms, and had duplicated sequences removed. 
```{bash, eval=FALSE}
##this way seems to be the most concise
#cut out column 9, remove the first 4 lines, get only lines with protein_id,
#get the gene and protein ID
#
cut -f 9 aip_genomic_refseq.gtf | tail -n+5 | grep "protein_id" | \
awk -F';' '{for(i=1;i<=NF;i++) if ($i ~ /gene_id/ || $i ~ /protein_id/) printf $i; print ""}'| \ 
cut -d " " -f2,4 | sed 's/"//g' | uniq > ../02_output/04_aiptasia/aip_gene_protID.tab

#getting the protein lengths
seqtk comp aip_protein_refseq.faa | cut -f1,2 > ../02_output/04_aiptasia/aip_protein_lengths.tab
```

```{r}
#read in the data
aip_gene_protID <- read.table("../02_output/04_aiptasia/aip_gene_protID.tab", header = FALSE, sep = " ")
aip_gene_protID <- aip_gene_protID %>% 
  rename("gene_id" = V1, "protein_id" = V2)

aip_protein_lengths <- read.table("../02_output/04_aiptasia/aip_protein_lengths.tab", header = FALSE, sep = "\t")
aip_protein_lengths <- aip_protein_lengths %>% 
  rename("protein_id" = V1, "length" = V2)
```

## Merge the data
```{r, eval=FALSE}
#merge by protein_id  
aip_gene_prot_lengths <- aip_gene_protID %>% 
  left_join(aip_protein_lengths, by = "protein_id")
```

## checking amount of duplicates
```{r}
aip_longest_transcript_n_copies <- aip_gene_prot_lengths %>% 
  group_by(gene_id) %>% 
  filter(length == max(length)) %>% 
  summarize(n = n())
  
```
## taking a single duplicate per gene ID
```{r}
aip_long_transcript_distinct <- aip_gene_prot_lengths %>% 
  group_by(gene_id) %>% 
  filter(length == max(length)) %>% 
  distinct(gene_id, length, .keep_all = TRUE)

write.table(aip_long_transcript_distinct, "../02_output/04_aiptasia/aip_long_transcript_distinct.tab", sep = "\t", quote = FALSE, row.names = FALSE)
write.table(aip_long_transcript_distinct$protein_id, "../02_output/04_aiptasia/aip_long_transcript_distinct_protein.txt", sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)
```

```{bash, eval=FALSE}
seqtk subseq ../../00_data/aip_protein_refseq.faa aip_long_transcript_distinct_protein.txt > aip_longest_isoforms.fasta
```

## Nematostella
#### extract gene_ids for proteins from gtf file
take only column containing gene + Protein ID
remove header
filter for lines that contain protein_ids
take only those to entries per row

```{bash, eval=FALSE}
##approach using awk
$ cut -f 9 nvec_genomic_jaNemVect1_1.gtf | tail -n+5 | grep "protein_id" | awk -F';' '{for(i=1;i<=NF;i++) if ($i ~ /gene_id/ || $i ~ /protein_id/) printf $i";"; print ""}' | uniq | sed -e "s/;/\t/g" -e "s/\"//g" -e "s/protein_id//g" -e "s/gene_id//g" -e "s/\t$//g" -e "s/ //g" > ../02_output/03_nvec/nvec_gene_protID.tab

#get protein lengths
$ seqtk comp nvec_protein_jaNemVect1_1.faa | cut -f1,2 > ../02_output/03_nvec/nvec_protein_lengths.tab

```
### Read into R and rename
```{r, eval=FALSE}
#read in the data
nvec_gene_protID <- read.table("../02_output/03_nvec/nvec_gene_protID.tab", header = FALSE, sep = "\t")
###Solution
#this only keeps the first occurance of a given gene_id and length, the result has a unique gene_id and length and only 1 protein ID per gene
nvec_long_transcript_distinct <- nvec_gene_prot_lengths %>% 
  group_by(gene_id) %>% 
  filter(length == max(length))%>% 
  distinct(gene_id, length, .keep_all = TRUE)

#save this we dont care about marker genes for now
write.table(nvec_long_transcript_distinct, "../02_output/03_nvec/nvec_long_transcript_distinct.tab", quote = FALSE, row.names = FALSE, col.names = TRUE)
write.table(nvec_long_transcript_distinct$protein_id, "../02_output/03_nvec/nvec_long_transcript_distinct_protein.txt", quote = FALSE, row.names = FALSE, col.names = F)
```


###get longest isoforms as fasta file
its saved as nvec_longest_isoforms.fasta
```{bash, eval=FALSE}
$ seqtk subseq ../../00_data/nvec_protein_jaNemVect1_1.faa nvec_long_transcript_distinct_protein.txt > nvec_longest_isoform.fasta
```


## Xenia
Here protein lengths could directly be extracted from the fasta file

```{bash, eval=FALSE}
#xenia prot lengths
seqtk comp xenSp1.proteins_hu2020.fa | cut -f1,2 > /home/leon/Masterarbeit/Projects/filter_proteomes/02_output/xenia_protein_lengths.tab
```

### Import data

```{r}
#import lengths
xenia_protein_lengths <- read.table("/home/leon/Masterarbeit/Projects/filter_proteomes/02_output/xenia_protein_lengths.tab", header = FALSE, sep = "\t")

xenia_protein_lengths <- xenia_protein_lengths %>% 
  rename("protein" = V1, "length" = V2)

```
### add gene ID
```{r}
xenia_protein_lengths$gene <- xenia_protein_lengths$protein %>%
  substr(1,9)
```
### filter longest transcript
```{r}
xenia_longest_transcript <- xenia_protein_lengths %>% 
  group_by(gene) %>% 
  filter(length == max(length))

#export protein as list
write.table(xenia_longest_transcript$protein, "/home/leon/Masterarbeit/Projects/filter_proteomes/02_output/xenia_longest_transcript.txt", quote = FALSE, row.names = FALSE, col.names = FALSE)
```

### get longest isoforms as fasta file
its saved as xenia_longest_isoforms.fasta
```{bash, eval=FALSE}
seqtk subseq ../00_data/xenSp1.proteins_hu2020.fa xenia_longest_transcript.txt > xenia_longest_isoforms.fasta
```

## Stylophora

### this way seems to be the most concise
```{bash, eval=FALSE}
$ cut -f 9 nvec_genomic_jaNemVect1_1.gtf | tail -n+5 | grep "protein_id" | awk -F';' '{for(i=1;i<=NF;i++) if ($i ~ /gene_id/ || $i ~ /protein_id/) printf $i; print ""}' | cut -d " " -f2,4 | sed 's/"//g' | uniq > ../02_output/02_spis/spis_gene_protID.tab


#getting the protein lengths
$ seqtk comp spis_proteome_refseq_17.faa | cut -f1,2 > ../02_output/02_spis/spis_protein_lengths.tab
```

### Read into R and rename
```{r, eval=FALSE}
#read in the data
spis_gene_protID <- read.table("../02_output/02_spis/spis_gene_protID.tab", header = FALSE, sep = "\t")
spis_gene_protID <- spis_gene_protID %>% 
  rename("gene_id" = V1, "protein_id" = V2)

spis_protein_lengths <- read.table("../02_output/02_spis/spis_protein_lengths.tab", header = FALSE, sep = "\t")
spis_protein_lengths <- spis_protein_lengths %>% 
  rename("protein_id" = V1, "length" = V2)
```
### Merge the data
```{r, eval=FALSE}
#merge by protein_id  
spis_gene_prot_lengths <- spis_gene_protID %>% 
  left_join(spis_protein_lengths, by = "protein_id")
```

### Filter longest isoform
```{r, eval=FALSE}
#this has duplicate genes due to duplicate proteinIDs in genome
spis_longest_transcript_length <- spis_gene_prot_lengths %>% 
  group_by(gene_id) %>% 
  summarize(length = max(length))

###Solution
#this only keeps the first occurance of a given gene_id and length, the result has a unique gene_id and length and only 1 protein ID per gene
spis_long_transcript_distinct <- spis_gene_prot_lengths %>% 
  group_by(gene_id) %>% 
  filter(length == max(length))%>% 
  distinct(gene_id, length, .keep_all = TRUE)
```

### checking the duplicates, there are more than 1 protein IDs per unique protein
```{r, eval=FALSE}  
#unique longest transcripts
spis_longest_transcript_unique <- spis_gene_prot_lengths %>% 
  group_by(gene_id) %>% 
  filter(length == max(length)) #%>% 
  #summarize(n = n())

#this gives the number of duplicates
spis_n_protein_duplicates <- spis_gene_prot_lengths %>% 
  group_by(gene_id) %>% 
  filter(length == max(length)) %>% 
  summarize(n = n())
```

### look for marker genes in duplicates
```{r, eval=FALSE}
#get all proteinIDs corresponding to duplicate gene IDs
#use this to lookup marker genes
spis_genes_with_duplicate_proteinIDs <- spis_n_protein_duplicates %>% 
  filter(n > 1) %>%
  left_join(spis_gene_prot_lengths, by = "gene_id") %>%
  select(gene_id, protein_id)

#find wether duplicate genes are marker genes
spis_genes_with_duplicate_proteinIDs$is_marker <- spis_genes_with_duplicate_proteinIDs$protein_id %in% spis_marker$gene_id

#export list of ´duplicate genes that are marker genes
write.table(spis_genes_with_duplicate_proteinIDs[spis_genes_with_duplicate_proteinIDs$is_marker,], "../02_output/02_spis/spis_markers_with_duplicate_proteinIDs.txt", quote = FALSE, row.names = FALSE, col.names = TRUE)
```

## add proteins to longest transcript that are marker genes
```{r, eval=FALSE}
# add marker genes duplicates to longest transcript
#going from spis_long_transcript_distinct which has a unique gene_id and length and only 1 protein ID per gene, while removing an arbitrary protein ID
#replace the protein IDs with the marker gene protein IDs
spis_marker_duplicates <- spis_genes_with_duplicate_proteinIDs[spis_genes_with_duplicate_proteinIDs$is_marker,] %>% 
  select(gene_id, protein_id)

spis_longest_transcript_distinct_markers <- spis_long_transcript_distinct %>% 
  anti_join(spis_marker_duplicates, by = "gene_id") %>% 
  bind_rows(spis_marker_duplicates)

```


## Export list
```{r, eval=FALSE}
write.table(spis_longest_transcript_distinct_markers$protein_id, "../02_output/02_spis/spis_longest_transcript_distinct_markers_protein.txt", quote = FALSE, row.names = FALSE, col.names = FALSE)

write.table(spis_longest_transcript_distinct_markers, "../02_output/02_spis/spis_longest_transcript_distinct_markers.tsv", quote = FALSE, row.names = FALSE, col.names = FALSE)
```

#get longest isoforms as fasta file
its saved as spis_longest_isoforms.fasta
```{bash, eval=FALSE}
seqtk subseq ../../00_data/spis_proteome_refseq_17.faa spis_longest_transcript_distinct_markers_protein.txt > spis_longest_isoforms.fasta
```
